---
title: "splinemixmeta"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{splinemixmeta}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Package `splinemixmeta` provides some basic capabilities in non-parametric meta-regression. If one has responses `y`, each with a standard error, from previous analyses, and wants to regress those against study-level explanatory variables `x` without assuming linearity, `splinemixmeta` allows an arbitrary smooth relationship to be estimated as a smoothing spline. It is also possible to include other explanatory variables (assumed to have a linear relationship to `y`) and random effects groupings. Such an approach may be referred to as "non-parametric meta-analysis", "spline meta-regression", "meta-splines", or other similar terminology mash-ups.

`splinemixmeta` works by bringing together features of `mgcv` for building spline components and `mixmeta` for estimating general mixed-effects meta-analysis models. Specifically, the splines of `mgcv` can be used as random effects components, with the penalty matrix providing a correlation structure for spline coefficients, the unknown smoothness parameter replaced by an unknown variance parameter, and any unpenalized directions in the spline parameter space (e.g. spline parameters representing a linear relationship) moved into fixed effects. The resulting pieces are then set up as needed for a call to `mixmeta::mixmeta` for estimating the model. Variance parameters are always estimated by REML (restricted maximum likelihood).

There are some important limits on what will work. At the moment, only spline choices for which covariance matrices can be diagonalized are supported. One can have multiple such components. In practice, the only bivariate spline (i.e. not simply addition of two univariate splines) that could potentially work is `bs = 'tp'`, although currently there is an issue translating the pieces from `mgcv` to `splinemixmeta` for that case. It is possible that `bs = 'mrf'` also works, but it has not been tested. See `help(splinemixmeta)` for more about what is supported.

The estimation machinery is not particularly efficient and so may be notably slow for large models.

# Example

Here we give a simulated example. Say we have `n=50` values, with 5 values from each of 10 groups, with a sinusoidal relationship between `y` and `x` to provide a simple nonlinear example. The groups will be treated as random effects, as will the spline.

## Data simulation

```{r fig.height = 4, fig.width = 6}
set.seed(1)
n <- 50
num_groups <- 10
group <- rep(1:num_groups, each = 5) |> factor()
group_effects <- rnorm(length(group), 0, 0.3)
x <- runif(n, 0, 20)
coef_x <- 0.6
se <- runif(n, 0.1, 0.3)
fx <- 1.5 * sin(2 * pi * x / 12)
y <- 10 + coef_x * x + fx + 
  group_effects[group] +
  fx + rnorm(n, 0, 1) + rnorm(n, 0, se)
data <- data.frame(y, x, se, group)
plot(y ~ x, data = data, col = group, pch = 19, main = "Simulated data (color = group)")
```

This data simulation has the following pieces:

- We have 10 groups of 5 data points each. Since the `x` values are drawn independently, the groups are randomly associated with `x` values.
- `x` is uniformly distributed from 0 to 20.
- Fixed effects for `y` include an intercept and linear term.
- Random effects for `y` include the group effects.
- The sinusoidal term for `y` will be estimated as a spline, treated as a random effect.
- Residuals are normally distributed with `sd = 1`.
- Measurement errors (i.e. estimation error from the previous studies from which `y` were obtained) are normally distributed with standard deviations that are the standard errors (`se`) from the previous studies. These are simulated as uniformly distributed between 0.1 and 0.3 to show heterogeneity in the precision of previous studies, which be due for example to different sample sizes.

## Estimation of a spline meta-regression model

The spline meta-regression model can be estimated like this:

```{r}
library(splinemixmeta)
# smm <- splinemixmeta( mgcv::s(x), y ~ x, se = se, manual_fixed = TRUE, data = data, random = ~ 1 | group)
smm <- splinemixmeta( mgcv::s(x, bs = "cr"), y ~ x, se = se, manual_fixed = TRUE, data = data, random = ~ 1 | group)
```

The object `smm` will have class "splinemixmeta" and "mixmeta". First we can look at a summary of the model:

```{r}
summary(smm)
```

The spline terms are labeled as `basisFxns_A`. Further splines would be `basixFxns_B`, etc. 

Two factors have been introduced. One is called `all`, which has a single level for every row of the data, which facilitates use of `mixmeta::mixmeta`. The other is called `ID`, which has a unique value for each row. In meta-regression, it can be tempting to think that the standard errors associated with `y` values serve as residual variation, but that is not typically the case. If each `y` came from a large sample size in the previous studies, the standard errors would be small, but we would still expect study-level variation around any regression line because the world is noisy. That study-level variation, i.e. residual variation, is set up as a random effect with one level for each `y`.

By using `y ~ x`, we included a linear term for `x` directly (manually), and thus we needed to set `manual_fixed = TRUE`. Otherwise `splinemixmeta` would have extracted a linear term from the spline setup and included it in the model, and we don't want it duplicated. In the `summary` output, we see the full covariance matrix from the spline random effect, which is the vector spline coefficients. This is shown as the standard deviation for each component (all the same) and their correlations (all 0), reflecting a covariance matrix that is a constant times an identity matrix.

Finally we see the estimated standard deviation between groups and the estimated residual standard deviation, shows as the `~1 | ID` term.

(If the default names `all` or `ID` are already used in the data set, `splinemixmeta` will choose different names.)

## Predictions

Predictions can be obtained at multiple levels of the random effects. The `predict` function for `splinemixmeta` objects simplifies this by allowing the simple choices of including any spline terms (default `TRUE`), other random effects (default `FALSE` but possibly of interest) and residuals (default `FALSE` and typically not of interest). Predictions will come with columns for standard errors and variances (simply the squared standard errors). The machinery for predictions is modified from `mixmeta::blup()`, where "blup" means "best linear unbiased predictor", a standard term in linear mixed effects modeling. Hence the column of predictions returned by `predict` is labeled "`blup`". The blups come from the conditional distribution of random effects, given the data and estimated variance parameters. Following `mixmeta::blup`, predictions can be of type "outcome", in which case variance from fixed effects terms is included in the prediction variance (and standard errors), or "residual", in which case only variance from random effects is included. The default is "outcome". See `help(predict.splinemixmeta)` for details. Any further fine-grained control (such as including one spline but not another), can be done by calling `splinemixmeta::blup()` directly.

Here are two versions of predictions:

```{r}
pred_spline_only <- predict(smm, include_smooths = TRUE, include_REs = FALSE, include_residuals = FALSE, type = "outcome")
pred_spline_and_groups <- predict(smm, include_smooths = TRUE, include_REs = TRUE, include_residuals = FALSE, type = "outcome")
message("Need to fix case with level=0")
# pred_fixed_only <- predict(smm, include_smooths = FALSE)
head(cbind(pred_spline_only, pred_spline_and_groups))
# blup(smm, level=0, vcov = TRUE, se = TRUE)
```
## Figures

`splinemixmeta` provides a basic capability for figures, for which the suggested package `ggplot2` is needed.

```{r fig.height = 4, fig.width = 6}
plot(smm, xvar = x, xlab = "x", ylab = "y", title = "Spline meta-regression fit")
```

